Useful biological models with no formal training

### 2013-11-26 Paper.tex
Changed sections headers and abstract to match Methodology paper format

\begin{fmbox}
\dochead{Software}
% \addtocontents{toc}{\protect\setcounter{tocdepth}{-1}}
% \firstpage{1}

\title[Bringing biological networks to life with ANIMO]{Bringing biological networks to life with ANIMO}

% \author[Schivo, Scholma \textit{et~al}]{Stefano Schivo\,$^{1,*}$, Jetse Scholma\,$^{2,*}$, Brend Wanders\,$^3$, Ricardo A. Urquidi Camacho\,$^2$, Paul E. van der Vet\,$^3$, Marcel Karperien\,$^2$, Rom Langerak\,$^1$, Jaco van de Pol\,$^{1,\ddag}$ and Janine N. Post\,$^{2,\ddag}$}
% \address{$^{1}$Formal Methods and Tools, Faculty of EEMCS, University of Twente, 7522NH Enschede, The~Netherlands;
% $^{2}$Developmental BioEngineering, MIRA Institute for Biomedical Technology and Technical Medicine, University of Twente, 7522NH Enschede, The~Netherlands;
% $^{3}$Human Media Interaction, Faculty of EEMCS, University of Twente, 7522NH Enschede, The~Netherlands}

@@

\def\dagger{AA}
\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   %corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={s.schivo@utwente.nl}   % email address
]{\inits{S}\fnm{Stefano} \snm{Schivo}}
\author[
   addressref={aff2},
   noteref={n1},
   email={j.scholma@utwente.nl}
]{\inits{J}\fnm{Jetse} \snm{Scholma}}
\author[
   addressref={aff3},
   email={b.wanders@utwente.nl}
]{\inits{B}\fnm{Brend} \snm{Wanders}}
\author[
   addressref={aff2},
   email={rauc6788@hotmail.com}
]{\inits{RA}\fnm{Ricardo A.} \snm{Urquidi Camacho}}
\author[
   addressref={aff3},
   email={p.e.vandervet@utwente.nl}
]{\inits{PE}\fnm{Paul E.} \snm{van der Vet}}
\author[
   addressref={aff2},
   email={h.b.j.karperien@utwente.nl}
]{\inits{M}\fnm{Marcel} \snm{Karperien}}
\author[
   addressref={aff1},
   email={r.langerak@utwente.nl}
]{\inits{R}\fnm{Rom} \snm{Langerak}}
\author[
   addressref={aff1},
   %corref={aff1},
   email={j.c.vandepol@utwente.nl}
]{\inits{J}\fnm{Jaco} \snm{van de Pol}}
\author[
   addressref={aff2},
   corref={aff2},
   email={j.n.post@utwente.nl}
]{\inits{JN}\fnm{Janine N.} \snm{Post}}

@@

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Motivation}
Computational modelling is a vital part of the empirical cycle,
formalizing existing knowledge and experimental data in a systematic way.
Existing methods are often parameter intensive, or supporting tools are
not accessible enough to the biological community. These factors
hamper widespread application of modelling in biological research.
\parttitle{Results}
We present an approach to modelling based on a series of
abstractions that limit the number of parameters in the model.
To support this approach, we developed a user-friendly software tool,
ANIMO (Analysis of Networks with Interactive MOdelling).
An ANIMO model of signal transduction events downstream of TNF$\alpha$
and EGF in HT-29 human colon carcinoma cells is described as a case study.
This model gives a formal description of crosstalk between the pathways at
different cellular levels and led to the formulation of novel testable hypotheses.
\parttitle{Availability and implementation}
ANIMO is implemented as a plug-in to the network visualization tool Cytoscape
and can be downloaded from \url{http://fmt.cs.utwente.nl/tools/animo}.
Additional information and a step-by-step installation guide can be found
in Supplementary Section~\ref{sec:animo-installation}. A user manual is 
provided in Supplementary Section~\ref{sec:animo-manual}.
\end{abstract}

@@

\begin{keyword}
\kwd{sample}
\kwd{article}
\kwd{author}
\end{keyword}

@@

\section*{Introduction}\label{sec:introduction}
In living cells, processes are regulated by networks of interacting molecules.
Aberrations in these networks underlie a wide range of pathologies. The development of new therapies
requires a thorough insight in the functioning of these networks. Obtaining such insight can be a challenging task.
Feedback loops and crosstalk between pathways lead to an intricate wiring of the network.
Hence, it is necessary to study the ensemble of molecules involved, because the 
behaviour of individual molecules is not sufficient for a complete understanding. 
Since the human brain is ill-suited to grasp the non-linear dynamics of these complex networks and
the entailed emergent properties, the role of computational support is increasing
in molecular biology.

The systems biology approach to understanding biological systems starts off from a
scientific question and then follows an empirical cycle \--\ or rather a positive spiral \--\ of
knowledge/theory $\rightarrow$ model $\rightarrow$ hypotheses $\rightarrow$ experiments $\rightarrow$
observations $\rightarrow$ update and/or refinement of knowledge/theory,
until an answer to the original question is found (Figure~\ref{fig:empirical-spiral}).
A model plays a pivotal role in this cycle:
\begin{enumerate}
  \item to organize data and store knowledge,
  \item to structure reasoning and discussion
  \item to perform \emph{in silico} experiments and derive hypotheses.
\end{enumerate}
An \emph{in silico} model is always a simplified representation of biological reality and is never the 
aim in itself.
Rather, it is a powerful means in the process of gaining an understanding of the biological system.
Given its role in the empirical cycle, the process of modelling is especially effective
when applied by the experts with respect to a certain biological system. Biologists usually have a good sense 
of cause-and-effect relationships of molecular interactions. In addition, they are the most knowledgeable
on the network topology and the dynamics of the biological system they are studying.
Since they also benefit most from the generation of hypotheses and from an efficient experimental design, 
biologists would be the primary candidates to construct models of their research topic.

As models are a formalization of knowledge or theories, an underlying formalism is needed to express
this knowledge. Different formal methods have been successfully applied to construct representations
of biological systems. Among these methods are Boolean logic~\cite{boolean-networks-flower,boolean-networks2},
ordinary differential equations (ODEs, reviewed by~\cite[]{hidde-review}),
interacting state machines~\cite{interacting-sm1,interacting-sm2},
process calculi~\cite{blenx,bio-pepa}, Timed Automata~\cite{ta-siebert,bartocci-oscillators,
oded-ode-ta-discretization} and Petri nets~\cite{petri-nets,petri-nets2}.
Most of these formal methods have been implemented into software tools to aid the process
of modelling. Due to the lack of such a supporting tool, \tas\ have remained a less 
frequently applied method.

\tas\ have been developed to model the dynamic behaviour of systems with processes running in parallel~\cite{timed-automata-alur-dill}. 
As such, \tas\ have been applied in communication protocols and industrial control engineering~\cite{ta-audio-protocol,ta-wap-gateway,ta-ws-bap}. The 
parallels between these application areas and regulatory processes in cells have triggered the step towards 
their use in biology.
\cite{ta-siebert} use \tas\ to extend a classical modelling paradigm
~\cite{thomas-formalism}, allowing to add temporal dynamics to gene network models.
\cite{bartocci-oscillators} describe a model of biological oscillators and test 
synchronization properties in this dynamic system.
A discretization of ODEs to \tas\ is proposed by~\cite{oded-ode-ta-discretization}, applying
a translation between the two formalisms to an example gene regulatory network. Two 
different approaches to transforming
a Petri net model into \tas\ are presented by~\cite{ta-giapponesi},
who also address the important issue of state space explosion in their paper.
Finally, \cite{ta-radiazioni} propose an \emph{ad hoc} \tas\ model of a radiation treatment
system, which is then validated through UPPAAL.\\
Each of these approaches has been successfully validated, demonstrating the potential of \tas\
in biological applications. However, these approaches were all limited to simple
or specific examples and none of these modelling methods
has led to a tool implementation of the proposed method to encourage a broader use
of \tas\ in molecular biology.

Mastery of most existing modelling tools requires training and experience in mathematical modelling. 
In this respect, a lack of tradition in quantitative
reasoning and formal methods within the biological community at large is still a stumbling block for
widespread application of modelling of biological systems. Here, we present an intuitive method for the
construction of formal in silico models of the dynamics of molecular networks, supported by a novel,
user friendly modelling tool, ANIMO (Analysis of Networks with Interactive 
MOdelling,~\cite[]{animo-bibe}). \tas\ are used as the underlying mathematical formalism.

In the Methods section, we will explain how choosing a suitable abstraction level can make the construction
of models more intuitive. We will then show how ANIMO is designed to support the modelling process following
this approach. Construction of a small model based on experimental data will exemplify the method that we
propose. In the Results section, we first show an ANIMO model of the genes and proteins that constitute
the circadian clock network in \emph{Drosophila Melanogaster}. The remainder of that section is dedicated to
illustrate how a single modelling iteration in the empirical cycle is used to compile prior knowledge and
experimental data into a model, perform \emph{in silico} simulations and derive meaningful testable 
hypotheses. These hypotheses are supported by literature on interactions in different cell types.

Timed Automata have been shown to be a powerful formalism to model biological processes
~\cite{ta-siebert,bartocci-oscillators,oded-ode-ta-discretization}. A timed automaton consists of locations
and transitions between these locations (see Fig.~\ref{fig:abstraction-mek-erk}), and a system of timed automata can be 
used to model a system of interacting molecules. At any time, each automaton is in a specific location, and together 
these locations represent the current state of the biological system. Each timed automaton can have one or more local clocks
associated to it, allowing temporal control of transitions between locations. These transitions are used to 
represent interactions between molecules. Fast interactions take less time than slow interactions 
to perform an activation or inhibition step. We have previously described in detail how the 
scenarios presented in Section~\ref{subsec:abstractions} can be used to calculate the timing of molecular 
interactions to give a description of network dynamics~\cite{animo-bibe}. Figure~\ref{fig:abstraction-mek-erk}
presents a small example that illustrates the basic properties of \tas. 
This model describes the activation of ERK by MEK\footnote{All acronyms used in this paper
and their corresponding UniProt IDs are listed in Suppl. Sect.~\ref{suppl-sec:names}.}.

\subsection*{ANIMO}
The modelling approach described in Section~\ref{subsec:abstractions} and Section~\ref{subsec:timed-automata}
is implemented in the
software tool ANIMO (Analysis of Networks with Interactive MOdelling, \cite{animo-bibe}) as
a plug-in to the network visualization tool Cytoscape~\cite{cytoscape}. The visual interface of Cytoscape
makes the construction, expansion and rewiring of a network topology a fast and user-friendly process.

\subsection*{Using ANIMO to build a model based on data}\label{subsec:case-study}
To illustrate the process of modelling with ANIMO, we show the construction of a small model based on a 
literature compendium of signal transduction events in HT-29 human colon carcinoma cells~\cite{pathway-compendium}. 
This data set comprises triplicate
measurements of 11 different protein activities or post-translational modification states at 13 time points after
treatment with different combinations of TNF$\alpha$, EGF and insulin.
The data set contains relative protein levels and activities, which is a typical situation in biochemistry.

As a first step, we normalized measurements for each protein to the
maximum value in the complete experiment. This normalization results in a nondimensionalized data set that 
is suitable for use with ANIMO.

In Figure~\ref{fig:small-model}, we show the stepwise construction of a model of a small part of the network that is
able to account for measured variations in activity of IKK, JNK1, MK2, Casp8 and Casp3 upon stimulation with 100~ng/ml TNF$\alpha$.
In this example we aimed for inclusion of a minimum number of nodes in the network, 
while preserving biological relationships.
Multi-step cascades were aggregated into a single step when possible. Parameters for all interactions were 
set manually, and a close match was obtained between the model and the patterns that are present in the dataset.

A more comprehensive model based on the same dataset is presented in Section~\ref{subsec:case-study-larger}.

\section*{Results}\label{sec:results}
\subsection*{Expressivity of ANIMO models}\label{sec:animo-drosophila}
Results obtained with ANIMO are comparable to results with other modeling
approaches. To demonstrate this, Figure~\ref{fig:drosophila-model}
represents an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster}, based on the work
by~\cite{drosophila-ode-model}, where ordinary differential equations (ODEs) were used.
The cyclic behavior of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO act as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

\subsection*{Case study: using ANIMO to generate hypotheses}\label{subsec:case-study-larger}
In order to validate our modeling approach,
we constructed a larger model of the signaling network downstream of TNF$\alpha$ and EGF, formalizing
the crosstalk that takes place between the pathways at different levels of cellular regulation.
We first modeled the two pathways in isolation~(Suppl. Figs.~\ref{fig:large-model-tnf}, \ref{fig:large-model-egf}),
using information on protein interactions from
the KEGG~\cite{kegg} and phosphosite~\cite{phosphosite} databases. These models were fitted to experimental data
from studies by~\cite{pathway-compendium} and~\cite{pathway-autocrine}.
We then merged the two pathways into a single model and added autocrine crosstalk between the pathways that has been
described by~\cite{pathway-autocrine}.
Briefly, stimulation with TNF$\alpha$ leads to a rapid release of TGF$\alpha$ ({\sf TGFa} in the model),
which activates the EGF receptor ({\sf EGFR}).
This activation causes secretion of IL-1$\alpha$ ({\sf IL-1a}) at later time points.
The effect of IL-1$\alpha$ is down-regulated by the secretion of IL-1 receptor antagonist ({\sf IL-1ra})
downstream of TNF$\alpha$.
The resulting model (Fig.~\ref{fig:large-model-no-hypotheses}) was compared to the experimental data
for treatments with 100 ng/ml TNF alone and 100 ng/ml EGF alone (data not shown)~\cite{pathway-compendium}.

@@

@@

% \bibliographystyle{bmc-mathphys}
% \bibliography{Paper}

\begin{thebibliography}{}

\bibitem[Aldridge et~al., 2009]{pathway-fuzzy}
Aldridge, B.~B., Saez-Rodriguez, J., Muhlich, J.~L., Sorger, P.~K., and
  Lauffenburger, D.~A. (2009).
\newblock {Fuzzy logic analysis of kinase pathway crosstalk in
  TNF/EGF/Insulin-induced signaling}.
\newblock {\em PLoS Comput Biol}, 5(4):e1000340+.

\bibitem[Alur and Dill, 1994]{timed-automata-alur-dill}
Alur, R. and Dill, D.~L. (1994).
\newblock A theory of timed automata.
\newblock {\em Theor. Comput. Sci.}, 126(2):183--235.

\bibitem[Bailly et~al., 1996]{ap1-il1a}
Bailly, S., Fay, M., Isra{\"{e}}l, N., and Gougerot-Pocidalo, M.~A. (1996).
\newblock The transcription factor {AP-1} binds to the human interleukin 1
  alpha promoter.
\newblock {\em European Cytokine Network}, 7(2):125--128.

\bibitem[Bannister et~al., 1994]{cfos-cjun}
Bannister, A.~J., Brown, H.~J., Sutherland, J.~A., and Kouzarides, T. (1994).
\newblock {Phosphorylation of the c-Fos and c-Jun HOB1 motif stimulates its
  activation capacity.}
\newblock {\em Nucleic Acids Res}, 22(24):5173--6+.

\bibitem[Bartocci et~al., 2009]{bartocci-oscillators}
Bartocci, E., Corradini, F., Merelli, E., and Tesei, L. (2009).
\newblock Model checking biological oscillators.
\newblock {\em Electronic Notes in Theoretical Computer Science},
  229(1):41--58.
\newblock Proceedings of the Second Workshop From Biology to Concurrency and
  Back (FBTC 2008).

\bibitem[Batt et~al., 2007]{oded-ode-ta-discretization}
Batt, G., Salah, R.~B., and Maler, O. (2007).
\newblock On timed models of gene networks.
\newblock In {\em Proceedings of the 5th international conference on Formal
  modeling and analysis of timed systems}, FORMATS'07, pages 38--52, Berlin,
  Heidelberg. Springer-Verlag.

\bibitem[Bengtsson et~al., 2002]{ta-audio-protocol}
Bengtsson, J., Griffioen, W.~D., Kristoffersen, K.~J., Larsen, K.~G., Larsson,
  F., Pettersson, P., and Yi, W. (2002).
\newblock Automated verification of an audio-control protocol using {UPPAAL}.
\newblock {\em The Journal of Logic and Algebraic Programming}, 52-53(0):163 --
  181.

\bibitem[Bonzanni et~al., 2009]{petri-nets2}
Bonzanni, N., Krepska, E., Feenstra, K.~A., Fokkink, W., Kielmann, T., Bal,
  H.~E., and Heringa, J. (2009).
\newblock Executing multicellular differentiation: quantitative predictive
  modelling of c.elegans vulval development.
\newblock {\em Bioinformatics/computer Applications in The Biosciences},
  25:2049--2056.

\bibitem[Ciocchetta and Hillston, 2009]{bio-pepa}
Ciocchetta, F. and Hillston, J. (2009).
\newblock {Bio-PEPA: A framework for the modelling and analysis of biological
  systems}.
\newblock {\em Theor. Comput. Sci.}, 410:3065--3084.

\bibitem[Davis, 2000]{jnk-signaling}
Davis, R.~J. (2000).
\newblock Signal transduction by the {JNK} group of {MAP} kinases.
\newblock {\em Cell}, 103(2):239--252.

\bibitem[{de Jong}, 2002]{hidde-review}
{de Jong}, H. (2002).
\newblock Modeling and simulation of genetic regulatory systems: A literature
  review.
\newblock {\em Journal of Computational Biology}, 9:67--103.

\bibitem[Dematt\'{e} et~al., 2008]{blenx}
Dematt\'{e}, L., Priami, C., and Romanel, A. (2008).
\newblock {Modelling and simulation of biological processes in BlenX}.
\newblock {\em SIGMETRICS Perform. Eval. Rev.}, 35:32--39.

\bibitem[D{\'{e}}rijard et~al., 1994]{ras-jnk1}
D{\'{e}}rijard, B., Hibi, M., Wu, I.-H., Barrett, T., Su, B., Deng, T., Karin,
  M., and Davis, R.~J. (1994).
\newblock {JNK1}: A protein kinase stimulated by uv light and ha-ras that binds
  and phosphorylates the c-{Jun} activation domain.
\newblock {\em Cell}, 76(6):1025--1037.

\bibitem[Efroni et~al., 2003]{interacting-sm2}
Efroni, S., Harel, D., and Cohen, I.~R. (2003).
\newblock Toward rigorous comprehension of biological complexity: Modeling,
  execution, and visualization of thymic t-cell maturation.
\newblock {\em Genome Research}, 13(11):2485--2497.

\bibitem[Fisher et~al., 2005]{interacting-sm1}
Fisher, J., Piterman, N., Hubbard, E. J.~A., Stern, M.~J., and Harel, D.
  (2005).
\newblock Computational insights into caenorhabditis elegans vulval
  development.
\newblock {\em Proceedings of the National Academy of Sciences of the United
  States of America}, 102(6):1951--1956.

\bibitem[Gaudet et~al., 2005]{pathway-compendium}
Gaudet, S., Janes, K.~A., Albeck, J.~G., Pace, E.~A., Lauffenburger, D.~A., and
  Sorger, P.~K. (2005).
\newblock {A compendium of signals and responses triggered by prodeath and
  prosurvival cytokines.}
\newblock {\em Mol Cell Proteomics}, 4(10):1569--1590.

\bibitem[Hessel and Pettersson, 2007]{ta-wap-gateway}
Hessel, A. and Pettersson, P. (2007).
\newblock Model-based testing of a wap gateway: an industrial case-study.
\newblock In {\em Proceedings of the 11th international workshop, FMICS 2006
  and 5th international workshop, PDMC conference on Formal methods:
  Applications and technology}, FMICS'06/PDMC'06, pages 116--131, Berlin,
  Heidelberg. Springer-Verlag.

\bibitem[Hornbeck et~al., 2004]{phosphosite}
Hornbeck, P.~V., Chabra, I., Kornhauser, J.~M., Skrzypek, E., and Zhang, B.
  (2004).
\newblock {PhosphoSite}: A bioinformatics resource dedicated to physiological
  protein phosphorylation.
\newblock {\em Proteomics}, 4(6):1551--1561.

\bibitem[Janes et~al., 2005]{pathway-leastsquare}
Janes, K.~A., Albeck, J.~G., Gaudet, S., Sorger, P.~K., Lauffenburger, D.~A.,
  and Yaffe, M.~B. (2005).
\newblock A systems model of signaling identifies a molecular basis set for
  cytokine-induced apoptosis.
\newblock {\em Science}, 310(5754):1646--1653.

\bibitem[Kanehisa and Goto, 2000]{kegg}
Kanehisa, M. and Goto, S. (2000).
\newblock {KEGG: Kyoto Encyclopedia of Genes and Genomes}.
\newblock {\em Nucleic Acids Research}, 28(1):27--30.

>>>>

\begin{fmbox}
\dochead{Methodology}

\title{Useful biological models with no formal training}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   %corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={s.schivo@utwente.nl}   % email address
]{\inits{S}\fnm{Stefano} \snm{Schivo}}
\author[
   addressref={aff2},
   noteref={n1},
   email={j.scholma@utwente.nl}
]{\inits{J}\fnm{Jetse} \snm{Scholma}}
\author[
   addressref={aff3},
   email={b.wanders@utwente.nl}
]{\inits{B}\fnm{Brend} \snm{Wanders}}
\author[
   addressref={aff2},
   email={rauc6788@hotmail.com}
]{\inits{RA}\fnm{Ricardo A.} \snm{Urquidi Camacho}}
\author[
   addressref={aff3},
   email={p.e.vandervet@utwente.nl}
]{\inits{PE}\fnm{Paul E.} \snm{van der Vet}}
\author[
   addressref={aff2},
   email={h.b.j.karperien@utwente.nl}
]{\inits{M}\fnm{Marcel} \snm{Karperien}}
\author[
   addressref={aff1},
   email={r.langerak@utwente.nl}
]{\inits{R}\fnm{Rom} \snm{Langerak}}
\author[
   addressref={aff1},
   %corref={aff1},
   email={j.c.vandepol@utwente.nl}
]{\inits{J}\fnm{Jaco} \snm{van de Pol}}
\author[
   addressref={aff2},
   corref={aff2},
   email={j.n.post@utwente.nl}
]{\inits{JN}\fnm{Janine N.} \snm{Post}}

@@

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
The large amount of data with which an experimental biologist
must deal makes it largely impossible to reason on
the dynamics of complex systems without relying on computational
support.
In order to be fully profitable, computational modelling tools
must be both accessible and powerful, providing a means to
formalise and analyse experimental outcomes without requiring
additional training.
\parttitle{Results}
We show that the software tool ANIMO (Analysis of Networks with Interactive MOdelling) can be used
in different modelling tasks, and illustrate how this can be an asset for the biological researcher.
ANIMO models for \emph{Drosophila melanogaster} circadian clock and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells are used as case studies to illustrate the
possibilities offered by the tool.
\parttitle{Conclusions}
The software tool ANIMO allows the biologist to couple the
power of a formalism with the usability
of a user friendly interface, achieving useful insight without
the need to expand their knowledge of mathematical modelling tools.
\end{abstract}

@@

\begin{keyword}
\kwd{modelling}
\kwd{signalling pathway}
\kwd{timed automata}
\kwd{dynamic behaviour}
\end{keyword}

@@

\section*{Background}\label{sec:introduction}
In living cells, processes are regulated by networks of interacting molecules.
Aberrations in these networks underlie a wide range of pathologies. The development of new therapies
requires a thorough insight in the functioning of these networks. Obtaining such insight can be a challenging task.
Feedback loops and crosstalk between pathways lead to an intricate wiring of the network.
Hence, it is necessary to study the ensemble of molecules involved, because the 
behaviour of individual molecules is not sufficient for a complete understanding. 
Since the human brain is ill-suited to grasp the non-linear dynamics of these complex networks and
the entailed emergent properties, the role of computational support is increasing
in molecular biology.

% The systems biology approach to understanding biological systems starts off from a
% scientific question and then follows an empirical cycle \--\ or rather a positive spiral \--\ of
% knowledge/theory $\rightarrow$ model $\rightarrow$ hypotheses $\rightarrow$ experiments $\rightarrow$
% observations $\rightarrow$ update and/or refinement of knowledge/theory,
% until an answer to the original question is found (Figure~\ref{fig:empirical-spiral}).
% A model plays a pivotal role in this cycle:
% \begin{enumerate}
%   \item to organize data and store knowledge,
%   \item to structure reasoning and discussion
%   \item to perform \emph{in silico} experiments and derive hypotheses.
% \end{enumerate}
% An \emph{in silico} model is always a simplified representation of biological reality and is never the 
% aim in itself.
% Rather, it is a powerful means in the process of gaining an understanding of the biological system.
% Given its role in the empirical cycle, the process of modelling is especially effective
% when applied by the experts with respect to a certain biological system. Biologists usually have a good sense 
% of cause-and-effect relationships of molecular interactions. In addition, they are the most knowledgeable
% on the network topology and the dynamics of the biological system they are studying.
% Since they also benefit most from the generation of hypotheses and from an efficient experimental design, 
% biologists would be the primary candidates to construct models of their research topic.

As models are a formalization of knowledge or theories, an underlying formalism is needed to express
this knowledge. Different formal methods have been successfully applied to construct representations
of biological systems. Among these methods are Boolean logic~\cite{boolean-networks-flower,boolean-networks2},
ordinary differential equations (ODEs, reviewed by~\cite{hidde-review}),
interacting state machines~\cite{interacting-sm1,interacting-sm2},
process calculi~\cite{blenx,bio-pepa}, Timed Automata~\cite{ta-siebert,bartocci-oscillators,
oded-ode-ta-discretization} and Petri nets~\cite{petri-nets,petri-nets2}.
Most of these formal methods have been implemented into software tools to aid the process
of modelling. Due to the lack of such a supporting tool, \tas\ have remained a less 
frequently applied method.

\tas\ have been developed to model the dynamic behaviour of systems with processes running in parallel~\cite{timed-automata-alur-dill}. 
As such, \tas\ have been applied in communication protocols and industrial control engineering~\cite{ta-audio-protocol,ta-wap-gateway,ta-ws-bap}. The 
parallels between these application areas and regulatory processes in cells have triggered the step towards 
their use in biology.
\cite{ta-siebert} use \tas\ to extend a classical modelling paradigm~\cite{thomas-formalism},
allowing to add temporal dynamics to gene network models.
\cite{bartocci-oscillators} describe a model of biological oscillators and test 
synchronization properties in this dynamic system.
A discretization of ODEs to \tas\ is proposed by~\cite{oded-ode-ta-discretization}, applying
a translation between the two formalisms to an example gene regulatory network. Two 
different approaches to transforming
a Petri net model into \tas\ are presented by~\cite{ta-giapponesi},
who also address the important issue of state space explosion in their paper.
Finally, \cite{ta-radiazioni} propose an \emph{ad hoc} \tas\ model of a radiation treatment
system, which is then validated through the analysis tool UPPAAL~\cite{uppaal}.\\
Each of these approaches has been successfully validated, demonstrating the potential of \tas\
in biological applications. However, these approaches were all limited to simple
or specific examples and none of these modelling methods
has led to a tool implementation of the proposed method to encourage a broader use
of \tas\ in molecular biology. We have recently introduced the software tool ANIMO
(Analysis of Networks with Interactive MOdelling,~\cite{animo-ieee}), which aims
at making available the power of \tas\ to the experimental biologists without requiring
them to be fluent in any mathematical formalism.

% Mastery of most existing modelling tools requires training and experience in mathematical modelling. 
% In this respect, a lack of tradition in quantitative
% reasoning and formal methods within the biological community at large is still a stumbling block for
% widespread application of modelling of biological systems. Here, we present an intuitive method for the
% construction of formal in silico models of the dynamics of molecular networks, supported by a novel,
% user friendly modelling tool, ANIMO (Analysis of Networks with Interactive 
% MOdelling,~\cite{animo-ieee}). \tas\ are used as the underlying mathematical formalism.

% In the Methods section, we will explain how choosing a suitable abstraction level can make the construction
% of models more intuitive. We will then show how ANIMO is designed to support the modelling process following
% this approach. Construction of a small model based on experimental data will exemplify the method that we
% propose. In the Results section, we first show an ANIMO model of the genes and proteins that constitute
% the circadian clock network in \emph{Drosophila Melanogaster}. The remainder of that section is dedicated to
% illustrate how a single modelling iteration in the empirical cycle is used to compile prior knowledge and
% experimental data into a model, perform \emph{in silico} simulations and derive meaningful testable 
% hypotheses. These hypotheses are supported by literature on interactions in different cell types.

In this paper we present ANIMO 2.0, an updated version of ANIMO which allows for a considerably
more efficient analysis of models. ANIMO 2.0 also introduces new features, such as the possibility to interrogate
an ANIMO model with model checking queries represented as human language sentences instead of
mathematical formulae. The case studies presented in the following sections are centered around possible usage scenarios
for ANIMO, and aim at illustrating how a biologist can profit from the tool in its latest iteration.
An installation manual for ANIMO 2.0 is available at \url{http://fmt.cs.utwente.nl/tools/animo/manual.html}.

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation}\label{sec:animo-drosophila}
Results obtained with ANIMO are comparable to results with other modeling
approaches. To demonstrate this, Figure~\ref{fig:drosophila-model}
represents an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster}, based on the work
by~\cite{drosophila-ode-model}, where ordinary differential equations (ODEs) were used.
The cyclic behavior of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO act as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

\subsection*{Exploiting formal methods: model checking}\label{subsec:model-checking}
To show why one should use a formal model like \tas, we demonstrate the application
one of the biggest advantages of \tas: state space-based analyses such as
model checking~\cite{model-checking} provided by UPPAAL~\cite{uppaal}.

The technique of model checking consists in verifying the truth of some properties
that describe the behaviour of a model. This is usually done by exploring every
possible evolution of the model starting from an initial state. The set of all states
that can be reached by a model is called \emph{state space}, and formulae describing
the evolution of a system are usually described by (a subset of) computation tree
logic (CTL,~\cite{ctl}).

\subsection*{Using ANIMO to generate hypotheses}\label{subsec:case-study-larger}
In order to validate our modeling approach,
we constructed a larger model of the signaling network downstream of TNF$\alpha$ and EGF, formalizing
the crosstalk that takes place between the pathways at different levels of cellular regulation.
We first modeled the two pathways in isolation~(Suppl. Figs.~\ref{fig:large-model-tnf}, \ref{fig:large-model-egf}),
using information on protein interactions from
the KEGG~\cite{kegg} and phosphosite~\cite{phosphosite} databases. These models were fitted to experimental data
from studies by~\cite{pathway-compendium} and~\cite{pathway-autocrine}.
We then merged the two pathways into a single model and added autocrine crosstalk between the pathways that has been
described by~\cite{pathway-autocrine}.
Briefly, stimulation with TNF$\alpha$ leads to a rapid release of TGF$\alpha$ ({\sf TGFa} in the model),
which activates the EGF receptor ({\sf EGFR}).
This activation causes secretion of IL-1$\alpha$ ({\sf IL-1a}) at later time points.
The effect of IL-1$\alpha$ is down-regulated by the secretion of IL-1 receptor antagonist ({\sf IL-1ra})
downstream of TNF$\alpha$.
The resulting model (Fig.~\ref{fig:large-model-no-hypotheses}) was compared to the experimental data
for treatments with 100 ng/ml TNF alone and 100 ng/ml EGF alone (data not shown)~\cite{pathway-compendium}.

@@

\section*{Conclusions}

Timed Automata have been shown to be a powerful formalism to model biological processes
~\cite{ta-siebert,bartocci-oscillators,oded-ode-ta-discretization}. A timed automaton consists of locations
and transitions between these locations (see Fig.~\ref{fig:abstraction-mek-erk}), and a system of timed automata can be 
used to model a system of interacting molecules. At any time, each automaton is in a specific location, and together 
these locations represent the current state of the biological system. Each timed automaton can have one or more local clocks
associated to it, allowing temporal control of transitions between locations. These transitions are used to 
represent interactions between molecules. Fast interactions take less time than slow interactions 
to perform an activation or inhibition step. We have previously described in detail how the 
scenarios presented in Section~\ref{subsec:abstractions} can be used to calculate the timing of molecular 
interactions to give a description of network dynamics~\cite{animo-ieee}. Figure~\ref{fig:abstraction-mek-erk}
presents a small example that illustrates the basic properties of \tas. 
This model describes the activation of ERK by MEK\footnote{All acronyms used in this paper
and their corresponding UniProt IDs are listed in Suppl. Sect.~\ref{suppl-sec:names}.}.

\subsection*{ANIMO}
The modelling approach described in Section~\ref{subsec:abstractions} and Section~\ref{subsec:timed-automata}
is implemented in the
software tool ANIMO (Analysis of Networks with Interactive MOdelling, \cite{animo-ieee}) as
a plug-in to the network visualization tool Cytoscape~\cite{cytoscape}. The visual interface of Cytoscape
makes the construction, expansion and rewiring of a network topology a fast and user-friendly process.

% 
% \subsection*{Using ANIMO to build a model based on data}\label{subsec:case-study}
% To illustrate the process of modelling with ANIMO, we show the construction of a small model based on a 
% literature compendium of signal transduction events in HT-29 human colon carcinoma cells~\cite{pathway-compendium}. 
% This data set comprises triplicate
% measurements of 11 different protein activities or post-translational modification states at 13 time points after
% treatment with different combinations of TNF$\alpha$, EGF and insulin.
% The data set contains relative protein levels and activities, which is a typical situation in biochemistry.
% 
% As a first step, we normalized measurements for each protein to the
% maximum value in the complete experiment. This normalization results in a nondimensionalized data set that 
% is suitable for use with ANIMO.
% 
% In Figure~\ref{fig:small-model}, we show the stepwise construction of a model of a small part of the network that is
% able to account for measured variations in activity of IKK, JNK1, MK2, Casp8 and Casp3 upon stimulation with 100~ng/ml TNF$\alpha$.
% In this example we aimed for inclusion of a minimum number of nodes in the network, 
% while preserving biological relationships.
% Multi-step cascades were aggregated into a single step when possible. Parameters for all interactions were 
% set manually, and a close match was obtained between the model and the patterns that are present in the dataset.
% 
% A more comprehensive model based on the same dataset is presented in Section~\ref{subsec:case-study-larger}.

@@

\bibliographystyle{bmc-mathphys}
\bibliography{Paper}
